Dessicated diversification: how aridification slowed down speciation of Australian blind snakes

### 2023-01-21 manuscript.Rmd
update needs more work on the aridification/diversification paragraph

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo and mode with which different groups diversified. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annums) with the development of pockets of aridity [@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, arid-adapted vegetation expanded and mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to have promoted the rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;
@owenHow2017].

@@

To estimate paleoclimate in Australia from 40 Ma to the present-day, we used three two alternative approaches. The first approach used information derived from paleo-Köppen climate bands from @scoteseAtlas2021. Köppen bands are estimated from different fossil and lithological indicators and give a broad categorisation of spatial variation in major climate types (tropical, sub-tropical arid, warm temperate, cool temperate, and polar). We obtained estimated Köppen climate bands at 5 Ma intervals from 40 Ma to 0 Ma from and used the sub-tropical arid Köppen band as a binary aridity index. We removed sites found in the ocean by using digital elevation models (DEMs) to mask sites below sea-level. To account for uncertainty in paleogeographic reconstructions of continental movements and sea-levels, we used two alterative DEMs [@straumeGlobSed2019;@scotesePALEOMAP2018] to mask the ocean sites and provide two estimates of paleoaridity. To estimate paleotemperature from the Köppen bands, we followed the method of @scoteseAtlas2021 as implemented by @hagenEarth2021. Here, each Köppen climate band was assigned a temperature value based on present-day averages of those climate bands (tropical = 26, sub-tropical arid = 22, warm temperate = 16, cool temperate = 5, polar = -20). To account for global fluctuations in temperature, we then modified these values by a global temperature curve estimate from oxygen isotope data (66). Finally, to account for change of temperature with elevation, we used a lapse rate of 9.8°C/km. Following @hagenEarth2021, we model variation in the intensity of temperature oscillations during the Pleistocene. We also used the two alternative DEMs to give two alternative temperature curves using this approach [@straumeGlobSed2019;@scotesePALEOMAP2018].

@@

Our best-fitting model show that aridity is negatively associated with speciation, where speciation decreased as aridity increased. Two of our reconstructed paleoaridity index data show that aridity increased from ~25 Ma following the opening of the Southern Ocean and the onset of the Antartic circumpolar current [@byrneBirth2008]. A particularly sharp increase from 10 Ma followed from the "Hill gap" period of cooling [@hillHistory1994], and intensification from 5 Ma onwards following the Pliocene transition to a contemporary arid Australian climate [@byrneBirth2008]. Out of the 38 lineages we considered, 5/36 divergence events occurred in the last 10 million years when the continent intensively aridified. In contrast, 31/36 divergence event occurred between 25 – 10 Ma when the continent was getting warming but not entirely arid. Indeed 12/16 lineages that have current distributions in the arid biome diverged before 10 Ma. This pattern is consistent with other studies on vertebrates that have shown that the divergence time for some arid-adapted taxa pre-dated the intense aridification in the last 5 million years [@byrneBirth2008;@catulloAridification2014].

@@

## Diversification in the arid zone

Several studies have shown that arid-adapted species originated either from a single ancestor present in Australia before the Miocene, or through multiple independent expansion into the arid zone [@byrneBirth2008]. Our ancestral biome analyses show that several non-arid ancestors of _Anilios_ blind snakes have diverged into the arid zone evident by multiple arid species throughout the phylogeny. This pattern is consistent with several other groups of reptiles that started to diversify in the Miocene including monitor lizards [@brennanPhylogenomics2021;@pavon-vazquezCompetition2022], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], elapid snakes [@sandersMolecular2008], pythons [@esquerrePhylogenomics2020;@esquerreEvolutionary2022] and several geckos [@ashmanDiversification2018;@fujitaDiversification2010;@oliverIndependent2014].

>>>>

The present-day distribution of extant species can provide clues about their biogeographic history and insight into the tempo and mode with which different groups diversified. After a long period of relatively stable climate and mesic habitats, the aridification of continental Australia began around 20 Ma (mega annums) with the development of pockets of aridity [@martinCenozoic2006;@byrneDecline2011]. Subsequent extreme warming and cooling periods in the last ~5 Ma creating what is now known as the "outback" [@byrneBirth2008;@pepperLife2021]. During this period, arid-adapted vegetation expanded and mesic biomes became more fragmented [@crispHow2013;@martinCenozoic2006]. This period of aridification is believed to have promoted the rapid speciation of many arid-adapted vertebrates [@raboskyExceptional2007;@pepperPalaeoclimate2011;@brennanMass2017] and the extinction of some mesic-adapted groups [@bryantLines2016;@cardilloPhylogeny2017;
@owenHow2017].

@@

To estimate paleoclimate in Australia from 40 Ma to the present-day, we used three two alternative approaches. The first approach used information derived from paleo-Köppen climate bands from @scoteseAtlas2021. Köppen bands are estimated from different fossil and lithological indicators and give a broad categorisation of spatial variation in major climate types (tropical, sub-tropical arid, warm temperate, cool temperate, and polar). We obtained estimated Köppen climate bands at 5 Ma intervals from 40 Ma to 0 Ma from and used the sub-tropical arid Köppen band as a binary aridity index. We removed sites found in the ocean by using digital elevation models (DEMs) to mask sites below sea-level. To account for uncertainty in paleogeographic reconstructions of continental movements and sea-levels, we used two alterative DEMs [@straumeGlobSed2019;@scotesePALEOMAP2018] to mask the ocean sites and provide two estimates of paleoaridity. To estimate paleotemperature from the Köppen bands, we followed the method of @scoteseAtlas2021 as implemented by @hagenEarth2021. Here, each Köppen climate band was assigned a temperature value based on present-day averages of those climate bands (tropical = 26, sub-tropical arid = 22, warm temperate = 16, cool temperate = 5, polar = -20). To account for global fluctuations in temperature, we then modified these values by a global temperature curve estimate from oxygen isotope data [@songSeawater2019]. Finally, to account for change of temperature with elevation, we used a lapse rate of 9.8°C/km. Following @hagenEarth2021, we model variation in the intensity of temperature oscillations during the Pleistocene. We also used the two alternative DEMs to give two alternative temperature curves using this approach [@straumeGlobSed2019;@scotesePALEOMAP2018].

@@

Our best-fitting model show that aridity is negatively associated with speciation, where speciation decreased as aridity increased. Two of our reconstructed paleoaridity index data show that aridity increased from ~25 Ma following the opening of the Southern Ocean and the onset of the Antartic circumpolar current [@byrneBirth2008]. A particularly sharp increase from 10 Ma followed from the "Hill gap" period of cooling [@hillHistory1994], and intensification from 5 Ma onwards following the Pliocene transition to a contemporary arid Australian climate [@byrneBirth2008]. Out of the 38 lineages we considered, 5/36 divergence events occurred in the last 10 Ma when the continent intensively aridified. In contrast, 31/36 divergence event occurred between 25 – 10 Ma when the continent was getting warmer but not entirely arid. Indeed 12/16 lineages that have current distributions in the arid biome diverged before 10 Ma. This pattern is consistent with other studies on vertebrates that have shown that the divergence time for some arid-adapted taxa pre-dated the intense aridification in the last 5 Ma [@byrneBirth2008;@catulloAridification2014].

The observed overall decrease in diversification rate with increase aridification may be associated with mass extinction. We hypothesise that mesic-adapted ancestral blind snakes may have once been widespread but populations later became extinct due to extreme aridity in the last 5 Ma [@fujiokaAustralian2009], which possibly made even subterranean conditions uninhabitable for blind snakes and their social insect prey. The surviving population then became isolated in small patches of "mesic refugia" as the arid zone expanded, as has been proposed in Australian cicadas [@owenHow2017]. Contemporary distribution of extant lineages supports this hypothesis. Occurrence data of the 13/38 arid lineages are usually found in pockets of climatically stable oasis rather than dry desolate sand dessert.

However, extinction events are difficult to identify when fossil data is lacking [@loucaWhy2021;@raboskyExtinction2010].

The surviving lineages in the arid zone are often associated with

 due to their small size and home range,

 as  when Australia

Although many lineages occur in what is now the arid zone,  surviving lineages occur in the arid zone 
For instance,

@@

## Diversification in the arid zone
Several studies have shown that arid-adapted species originated either from a single ancestor present in Australia before the Miocene, or through multiple independent expansion into the arid zone [@byrneBirth2008]. Our ancestral biome analyses show that several non-arid ancestors of _Anilios_ blind snakes have diverged into the arid zone evident by multiple arid species throughout the phylogeny. This pattern is consistent with several other groups of reptiles that started to diversify in the Miocene including monitor lizards [@brennanPhylogenomics2021;@pavon-vazquezCompetition2022], _Sphenomorphus_ and _Eugongylus_ skinks [@skinnerLygosomine2011], elapid snakes [@sandersMolecular2008], pythons [@esquerrePhylogenomics2020;@esquerreEvolutionary2022] and several geckos [@ashmanDiversification2018;@fujitaDiversification2010;@oliverIndependent2014].
